---
title: "Module 3-1-Principle - Data Visualization with ggplot2 in R"
author: "Abigail Sewell"
date: '`r format(Sys.time())`'
format: 
  html: 
    toc: true
    toc-depth: 4
    toc-location: right-body
    number-sections: false
    code-line-numbers: true
    code-fold: false
    code-link: true
    embed-resources: true
    df-print: paged
editor: visual
execute: 
  freeze: auto
  warning: false  
  error: true
editor_options:
  chunck_output_type: console
  chunk_output_type: console
---

# Overview

## Expected Learning Outcomes

> After taking this workshop, participants should be able to do following:

1.  Explain the concept of the grammar of graphics when visualizing data with the ggplot2 package.

2.  Be familiar with various types of charts.

3.  Visualize data in counts and proportions.

4.  Select appropriate charts based on strategic considerations (e.g., the characteristics of the data and audience).

5.  Create a chart that involves one or two variables with either categorical or continuous data.

6.  Create a chart by adding a categorical moderator (3rd variable) to the chart involving two or three variables.

7.  Create correlation charts.

8.  Read charts and generate insights.

9.  Describe three popular packages that allow one to visualize data.

10. Explain the concept of the grammar of graphics when visualizing data with the ggplot2 package.

## Loading Packages

```{r loading up packages}
#options(repos = c(CRAN = "https://cloud.r-project.org"))
#install.packages("ggplot2")
#install.packages('ggrepel')
#install.packages('ggthemes')
#install.packages('scales')
#install.packages('plotly')
#install.packages('lattice')
#install.packages('GGally')
#install.packages("dplyr")
#install.packages("tidyverse")
#install.packages('ggtext')
#install.packages("glue")
library(ggplot2) #visualization
library(ggrepel) #labels for data
library(ggthemes) #collections of themes
library(scales) # scale
library(plotly) # interactive chart
library(GGally) # correlation
library(dplyr) # data transformation
library(tidyverse) # mega package containing 8 packages
library(ggtext) # for text visualization
library(glue) # combining multiple component
library(gapminder)

```

# 1. Understand mtcars data

## 1.1 Using Help

```{r mtcars data}
?mtcars
```

> A data frame with 32 observations on 11 (numeric) variables.

-   \[, 1\] mpg Miles/(US) gallon

-   \[, 2\] cyl Number of cylinders

-   \[, 3\] disp Displacement (cu.in.)

-   \[, 4\] hp Gross horsepower

-   \[, 5\] drat Rear axle ratio

-   \[, 6\] wt Weight (1000 lbs)

-   \[, 7\] qsec 1/4 mile time

-   \[, 8\] vs Engine (0 = V-shaped, 1 = straight)

-   \[, 9\] am Transmission (0 = automatic, 1 = manual)

-   \[,10\] gear Number of forward gears

-   \[,11\] carb Number of carburetors Note\]

## 1.2 Reading data and converting to a tibble (cars)

-   short cut for adding r code chunk: Ctrl+alt+Insert\
-   short cut for executing a code ("running a code") is ctrl+enter

```{r}
mtcars
class(mtcars)

cars <- mtcars |>
  rownames_to_column() |>
  as_tibble() |>
  rename(model = rowname) |>
  print(n = 20, width = Inf)

cars
```

## 1.3 Simple Descriptive Statistics

```{r}
cars |>
  summary()

glimpse(cars) # base R 

# tidyverse
cars |>
  glimpse()
skimr::skim(cars)
```

# 2. Basic Plotting Methods in Base R

```{r}
# using a built in plotting function

hist(cars$disp, breaks = 10)

# barplot will be discussed later

```

# 3. Lattice package

```{r}
#install.packages("lattice")
library(lattice)
xyplot(mpg ~ wt, cars) #xyplot(y~x,data)

histogram(cars$disp, breaks = 10)
```

# 4. ggplot2

-   We will use ggplot 2 the best tool in the market for data vizuialzation from now on

## 4.1 Elaborate Example

### 4.1.1. Plotting

```{r}
#| label: complex
cars |>
  count(cyl)

easy_labels <- c("4" = "4 Cylinder Cars",
                 "6" = "6 Cylinder Cars",
                 "8" = "8 Cylinder Cars"
                 )

cars |>
  mutate(cyl = factor(cyl)) |>
  ggplot(aes(x = mpg, y = disp, color = cyl)) +
  geom_point(size = 3,
             color='black'
             ) + #geom
  # geom_jitter() +
   geom_smooth(method = lm) +
  # facet_grid(cols = vars(cyl),
   facet_wrap(~ cyl,
              #scales = "free_y",
              ncol = 1,
              strip.position = "top",
              labeller = labeller(cyl = easy_labels)
              ) + #faceting
  scale_y_continuous(limits = c(0, NA), expand = c(0,0)) +
  coord_flip() + #coordinate
  theme_economist() + #labels
  labs(title = 'MPG vs Displacement',
       x = 'Miles Per Gallon',
       y = 'Displacement') +
  theme(
    strip.placement = "outside",
    strip.background = element_blank(),
    panel.background = element_blank(),
    panel.grid = element_blank(),
    axis.line = element_line()
  ) +
  guides(color = 'none')
```

### 4.1.2 x and y are both continuous with moderator and as_labeller()

```{r}
easy_labels_n <- as_labeller(c(`4` = "4 Cylinder Cars",
                               `6` = "6 Cylinder Cars",
                               `8` = "8 Cylinder Cars"
                               )
                             )
ggplot(data = cars, aes(x = disp, y = mpg, color = factor(cyl))) + #data
  geom_point(size=3) + #geometry
  facet_grid(~ factor(cyl),
             labeller = easy_labels_n
             ) + #faceting
  theme_bw() + #theme type
  labs(title = 'MPG vs Displacement',
       x = 'Displacement',
       y = 'Miles Per Gallon',
       color = "# of Cylinder"
       ) +
  guides(color = 'none')

```
